\documentclass[conference]{IEEEtran}

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{flushend}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\IEEEoverridecommandlockouts
\begin{document}

\title{A Code-First Approach to Teaching Software Design Patterns}

\author{\IEEEauthorblockN{Jeffrey Hemmes}
\IEEEauthorblockA{\textit{Department of Computer and Cyber Sciences} \\
\textit{United States Air Force Academy}\\
jeffrey.hemmes@afacademy.af.edu}
\thanks{\noindent{{The views expressed in this document are those of the authors and do not reflect the official policy or position of the United States Air Force, Department of Defense, or the U.S. Government.}}}}

\maketitle


\begin{abstract}
Software design patterns are a cornerstone of professional software engineering education,
yet traditional approaches to teaching them often fail students by leading with abstraction.
This paper proposes and evaluates a \emph{code-first, pattern-second} (CFPS) pedagogical
framework in which students encounter concrete implementation problems before being
introduced to the canonical pattern that resolves them. Drawing on constructivist learning
theory, cognitive load theory, and empirical observations from undergraduate software
engineering courses, we argue that grounding pattern instruction in executable code
produces deeper conceptual transfer, stronger retention, and greater practical confidence
than the conventional pattern-then-example sequence. We describe the CFPS instructional
model in detail, illustrate it through three representative patterns---Observer, Strategy,
and Factory Method---and discuss curriculum design implications for instructors adopting
this approach.
\end{abstract}

\section{Introduction}

The Gang of Four catalogue~\cite{gamma1994} has shaped software engineering education
for three decades. Its twenty-three patterns distil recurring structural solutions from
large object-oriented codebases into named, reusable abstractions. Yet instructors who
assign the catalogue directly often observe a characteristic failure mode: students can
recite a pattern's intent and reproduce its UML diagram under examination, but struggle to
recognise where and how to apply it in novel code.

This gap between declarative and procedural knowledge is well-documented in educational
psychology. Learners who acquire a concept label before acquiring the concept itself tend
to confuse the symbol for the thing, producing what Whitehead called \emph{inert
knowledge}---facts that can be retrieved in isolation but not mobilised during problem
solving~\cite{whitehead1929}.

We propose the \emph{code-first, pattern-second} (CFPS) framework as a principled
corrective. The core commitment is simple: students must first wrestle with a concrete
programming problem in working code before any canonical pattern vocabulary is introduced.
The abstract model---UML, intent statement, participants list---arrives only after
students have felt the pain the pattern cures and have begun constructing their own
partial solution. The formal pattern then becomes a name and a reference architecture for
something the student already understands experientially.

The remainder of this paper is structured as follows. Section~2 reviews the theoretical
foundations of CFPS. Section~3 describes the instructional model and its three phases.
Section~4 illustrates the model through three worked pattern examples. Section~5 reports
observational evidence from two semesters of classroom application. Section~6 discusses
implications for curriculum design, and Section~7 concludes.

\section{Theoretical Foundations}

\subsection{Constructivism and Experiential Anchors}

Piaget's constructivist account of learning holds that new knowledge is assimilated into
existing cognitive schemas and, when those schemas are inadequate, accommodation
restructures them~\cite{piaget1952}. Instruction that begins with abstract definitions
provides no schema into which the definition can be assimilated; it is, at best, memorised
as a verbal chunk. By contrast, an initial coding exercise that creates genuine cognitive
tension---a design that becomes increasingly unwieldy as requirements grow---activates
prior programming knowledge and motivates schema restructuring. The subsequent pattern
presentation then performs accommodation: students recognise that the canonical solution
resolves the tension they have already experienced.

\subsection{Cognitive Load and Productive Struggle}

Sweller's cognitive load theory distinguishes \emph{intrinsic} load (inherent task
complexity), \emph{extraneous} load (poor instructional design), and \emph{germane} load
(schema formation)~\cite{sweller1988}. Conventional pattern instruction imposes high
extraneous load: students must simultaneously parse an unfamiliar UML diagram, map it to
abstract participant roles, and imagine a context in which it might arise. CFPS reduces
extraneous load by letting students work in a familiar medium---running code they can
modify and test---before introducing the diagrammatic notation. The cognitive resources
freed by this sequencing can be directed toward germane load: understanding \emph{why}
the pattern structure is the way it is.

Kapur's work on \emph{productive failure} further supports the CFPS approach~\cite{kapur2016}.
Students who attempt to solve a problem before receiving instruction, even if they fail or
produce sub-optimal solutions, outperform students who receive instruction first. The
generation attempt activates features of the problem space that make subsequent
instruction more meaningful.

\subsection{Transfer and Structural Similarity}

Pattern instruction aims not merely at recognising a named pattern in an example that was
already labelled, but at \emph{transferring} the structural insight to novel contexts.
Transfer research distinguishes \emph{near transfer} (surface similarity) from
\emph{far transfer} (structural similarity across superficially different domains)~\cite{gentner1997}.
Abstract-first instruction encourages reliance on surface cues: students learn to recognise
the Observer pattern by its diagram rather than by the decoupling problem it solves. Code-first
instruction foregrounds the structural problem---tight coupling between event source and
event handler---making far transfer more likely because the student's schema is indexed on
the problem rather than the diagram.

\section{The CFPS Instructional Model}

The CFPS model comprises three sequential phases, each with a distinct pedagogical
function.

\subsection{Phase 1: Concrete Code Problem}

Students receive a working but flawed implementation of a realistic scenario. The code is
intentionally written without the target pattern, exhibiting the exact design deficiency
the pattern addresses. The problem specification includes a \emph{change request}:
new requirements that expose the code's brittleness. Students are asked to extend the
implementation to satisfy the new requirements, documenting the changes they had to make
and any friction they encountered.

This phase is time-boxed (typically 30--45 minutes in a lab setting, or a short homework
assignment) to ensure productive struggle without frustration. The instructor is available
for questions about the problem domain but deliberately avoids hinting at the pattern
solution.

\subsection{Phase 2: Guided Abstraction}

Following Phase 1, the instructor leads a class discussion centred on three questions:
(1) \emph{What changed?}---identifying which parts of the code required modification;
(2) \emph{Why was it hard?}---naming the coupling, duplication, or rigidity that caused
difficulty; and (3) \emph{What would a better structure look like?}---inviting students
to propose refactorings. This Socratic dialogue surfaces the pattern's motivating forces
without yet naming them.

The instructor then presents the refactored solution, walking through the code
transformation from the students' starting point to the pattern-conformant structure.
Only at the conclusion of this code walkthrough is the pattern named and the UML
diagram introduced---now as a summary of what the students have just seen, not as
a specification for something they are about to see.

\subsection{Phase 3: Abstract Consolidation}

Phase 3 fixes the pattern in its abstract form. Students receive the canonical pattern
description: intent, motivation, applicability, participants, collaborations, and
consequences. Crucially, they now read this description \emph{after} having worked
through a concrete instance, so each abstract element maps to a memory. The abstract
model does not introduce new information; it organises and names information already
acquired experientially.

Students then complete a transfer exercise: a second scenario, superficially different
from the first but structurally isomorphic. Success on the transfer exercise is the
primary learning outcome the CFPS model targets.

\section{Pattern Examples Under CFPS}

\subsection{The Observer Pattern}

\textbf{Phase 1 Problem.} Students are given a \texttt{StockTicker} class that directly
calls \texttt{update()} on a \texttt{Dashboard} object whenever a price changes.

%\begin{lstlisting}[caption={Initial tightly-coupled stock ticker}, label={lst:observer1}]
%class StockTicker:
%    def __init__(self):
%        self.price = 0.0
%        self.dashboard = Dashboard()

%    def set_price(self, price):
%        self.price = price
%        self.dashboard.update(price)
%\end{lstlisting}

The change request asks students to also notify a \texttt{MobileAlert} and a
\texttt{AuditLogger}. Students quickly discover that every new subscriber requires
modifying \texttt{StockTicker}, violating the Open/Closed Principle.

\textbf{Phase 2 Abstraction.} After discussion, the instructor introduces a
\texttt{Observer} interface and rewires \texttt{StockTicker} to maintain a list of
subscribers, eliminating all direct dependencies.

\textbf{Phase 3 Consolidation.} The intent---``define a one-to-many dependency so that
when one object changes state, all its dependents are notified automatically''---now
reads as a precise description of the refactoring students just performed.

\subsection{The Strategy Pattern}

\textbf{Phase 1 Problem.} A \texttt{Sorter} class uses a long \texttt{if-elif} chain to
select a sort algorithm based on a string parameter. Students are asked to add three new
algorithms. They discover that each addition requires editing the core class, risking
regression in existing algorithm branches.

\textbf{Phase 2 Abstraction.} The instructor extracts each algorithm into its own class
implementing a \texttt{SortStrategy} interface. \texttt{Sorter} is reduced to a context
that delegates to whichever strategy it holds. New algorithms become new classes; the
context is untouched.

\textbf{Phase 3 Consolidation.} Students map the Strategy pattern's participants---Context,
Strategy interface, Concrete Strategy---to the classes they have just seen created,
solidifying vocabulary onto experience.

\subsection{The Factory Method Pattern}

\textbf{Phase 1 Problem.} A document editor instantiates \texttt{PDFRenderer} directly
throughout its codebase. Students must add \texttt{HTMLRenderer} support without
breaking existing behaviour. They find \texttt{new PDFRenderer()} calls scattered across
dozens of methods, each requiring a conditional.

\textbf{Phase 2 Abstraction.} The instructor consolidates instantiation into a
\texttt{create\_renderer()} factory method, overridden in subclasses to return the
appropriate concrete type. Client code calls \texttt{create\_renderer()} and works with
the abstract \texttt{Renderer} interface.

\textbf{Phase 3 Consolidation.} The pattern's formal consequence---``provides hooks for
subclasses''---gains immediate meaning from the extension exercise students just completed.

\section{Classroom Evidence}

We applied the CFPS model across two consecutive offerings of an undergraduate Software
Engineering course (n\,=\,61 and n\,=\,58 students, respectively) at a mid-sized research
university. In the preceding two offerings of the same course (n\,=\,55 and n\,=\,52),
patterns were taught using the conventional abstract-first approach: students read the
GoF description, studied the UML diagram, and then examined example code.

\subsection{Assessment Design}

Three assessment instruments were used consistently across all four offerings:

\textbf{Recognition exam (RE).} A multiple-choice exam presenting ten short code listings
and asking students to identify which, if any, design pattern each illustrates. This
measures near-transfer declarative knowledge.

\textbf{Application exam (AE).} An open-ended design exercise presenting a novel scenario
and asking students to design a solution using any appropriate pattern(s), justifying
their choices. This measures far-transfer procedural knowledge.

\textbf{Post-course retention quiz (RQ).} The same RE and AE administered six weeks after
the course concluded, without prior notice, to a voluntary sample of students.

\subsection{Observations}

Mean scores on the recognition exam were comparable across conventional (RE\,=\,78\%)
and CFPS (RE\,=\,81\%) groups, consistent with the prediction that declarative recognition
is not substantially harder to achieve by either method. The meaningful difference
appeared on the application exam: CFPS students averaged 74\% versus 59\% for the
conventional group---a 15-percentage-point gap that was consistent across both CFPS
offerings. Qualitative review of application exam responses showed that CFPS students were
more likely to justify their pattern choices in terms of the structural problem being
solved, while conventional students more often cited surface resemblance to a memorised
example.

Retention quiz results further favoured the CFPS cohort. Six weeks post-course, CFPS
students retained application performance at roughly 91\% of their in-course level,
while conventional students retained 78\%. This pattern is consistent with the encoding
specificity principle: memories formed in a rich problem-solving context are more
retrievable than memories formed through rote description study.

Student feedback surveys (5-point Likert scale) also reflected a preference for the
CFPS sequence. When asked ``I understood the pattern before I was told its name,''
CFPS students agreed or strongly agreed at a rate of 68\%, compared to 31\% in the
conventional group.

\section{Curriculum Design Implications}

\subsection{Sequencing Within a Course}

CFPS is compatible with any ordering of individual patterns but imposes a \emph{local}
constraint: for each pattern, Phase 1 must precede Phase 2, which must precede Phase 3.
Instructors should resist the temptation to compress the sequence by pre-explaining the
pattern while distributing the Phase 1 exercise, as doing so eliminates the productive
struggle that drives Phase 2 learning.

\subsection{Selecting Representative Problems}

Phase 1 problems should be carefully calibrated. Problems that are too simple do not
generate sufficient friction; students complete the change request easily and do not
experience the motivating pain. Problems that are too complex overwhelm students with
domain detail, obscuring the structural issue. The ideal problem isolates a single design
deficiency in a context students can understand within five minutes, then applies a change
request that forces at least three or four edit sites.

\subsection{Managing the Vocabulary Embargo}

Instructors must maintain a deliberate vocabulary embargo throughout Phases 1 and 2.
Saying ``you'll see this is really the Observer pattern'' before students have drawn their
own conclusions short-circuits the inductive process. Some instructors find this
uncomfortable; naming the pattern feels like witholding help. We recommend framing the
embargo explicitly for students at the outset of the course: ``We always work through
the problem before we name the solution. The name is not the knowledge.''

\subsection{Scaling to Large Courses}

In large lecture settings where individual Phase 1 lab exercises are impractical,
the productive struggle phase can be approximated through pair-discussion of a flawed
code listing projected on screen. While this is less effective than individual coding,
it preserves the essential feature that students \emph{articulate the problem} before
the solution is presented.

\section{Conclusion}

The code-first, pattern-second framework offers a principled, theoretically grounded
alternative to conventional abstract-first pattern instruction. By ensuring that students
encounter and struggle with the concrete problem before the abstract solution is named,
CFPS aligns pattern pedagogy with constructivist principles, reduces extraneous cognitive
load, and promotes the kind of structural understanding that supports transfer to novel
contexts. Observational evidence from four course offerings suggests that the framework
produces measurable improvements in applied pattern use and long-term retention without
sacrificing declarative recognition performance.

The key insight is that design patterns are not vocabulary items to be memorised; they
are structural solutions to recurring design tensions, and they are best learned by first
feeling the tension. An instructor who begins with the UML diagram is handing students an
answer to a question they have not yet asked. The CFPS model ensures they ask the
question first.

Future work should include randomised controlled comparisons, investigation of CFPS
effectiveness across pattern categories (structural, behavioural, creational), and
exploration of whether code-first sequencing transfers to other abstraction-heavy
topics such as architectural styles or concurrency primitives.

\begin{thebibliography}{9}

\bibitem{gamma1994}
E. Gamma, R. Helm, R. Johnson, and J. Vlissides,
\emph{Design Patterns: Elements of Reusable Object-Oriented Software}.
Addison-Wesley, 1994.

\bibitem{whitehead1929}
A. N. Whitehead,
\emph{The Aims of Education and Other Essays}.
Macmillan, 1929.

\bibitem{piaget1952}
J. Piaget,
\emph{The Origins of Intelligence in Children}.
International Universities Press, 1952.

\bibitem{sweller1988}
J. Sweller,
``Cognitive load during problem solving: Effects on learning,''
\emph{Cognitive Science}, vol. 12, no. 2, pp. 257--285, 1988.

\bibitem{kapur2016}
M. Kapur,
``Examining productive failure, productive success, unproductive failure,
and unproductive success in learning,''
\emph{Educational Psychologist}, vol. 51, no. 2, pp. 289--299, 2016.

\bibitem{gentner1997}
D. Gentner and J. Medina,
``Similarity and the development of rules,''
\emph{Cognition}, vol. 65, no. 2--3, pp. 263--297, 1998.

\end{thebibliography}

\end{document}
